---
title: "Computational Graphics with R and ggplot2"
author: "Jeliazko Jeliazkov"
date: "January 26, 2015"
output: html_document
runtime: shiny
---

This is my attempt of giving an online component to my presentation such that the audience can follow along.

### Plotting a Map of the USA, Filled by Murder Rates

This plot is surprisingly simple to create once you get past the data manipulation.

First, we need to load the prerequisite packages. This can be done with either of the following commands:

```{r Loading Data}
library(maps)
require(ggplot2)
```

When writing functions, the require command is prefered as it fails more gracefully than the library function.
If neither of these packages are installed on your machine, feel free to use ```install.packages("maps"")``` or ```install.packages("ggplot2")```. Yes. The quotes really do matter. Now, load the packages.

Let's get the tricky data manipulation out of the way. We have a data containing the coordinates of the state borders (coming from the maps package) such that we can plot them and "connect the dots". To create a data frame from the maps data, we use the ggplot2 command ```map_data```. 

```{r Loading Map Data}
states <- map_data("state")
head(states)
```

We also have USA arrests data (I believe this is standard to R). This comes in a data frame, so let's just store it in a renamed variable since we will manipulate it later on.

```{r Loading Arrest Data}
arrests <- USArrests
head(USArrests)
```

Here comes the data manipulation. First, our column names are in two different cases. Let's standardize everything to lowercase.
```{r Manipulating Arrest Data}
names(arrests) <- tolower(names(arrests))
head(arrests)
```

This part requires a bit of foresight. Notice how the row names of the arrest data are the states. If we could make a separate column for this, we could merge the two datasets by state name (I believe the technical term is inner join).
```{r More Manipulation}
arrests$region <- tolower(rownames(USArrests))
head(arrests)
```

```{r Merging Data}
choro <- merge(states,arrests,by="region")
head(choro)
```

The merged data frame must be re-ordered to ensure the map is draw properly (ie the points are connected in order). Then, we can plot in one simple line using the command ```qplot```, which I believe stands for quick plot. The full command ```ggplot``` will be examined later on. In the line below, the longitude and latitude are given as the x and y arguments, with the data coming from the object ```choro```, and the rest should be evident.

```{r Plotting Map, fig.width=9, fig.height=5}
choro <- choro[order(choro$order),]
qplot(long,lat,data=choro,group=region, fill=murder, geom="polygon") + theme_bw(base_size=20)
```

### Loading Data, Plotting Points

While the previous example is neat, a scientist may be more interested in plotting his own data. To do so, we must first import data in R. This can be done through any ```read.*``` command. For example, tab delimited data can be imported using ```my_data_frame <- read.table("some_file.txt")``` and a comma separated file can be read using ```my_data_frame <- read.csv("other_file.csv")```. There are other commands, but those two are the ones I use most commonly.

This part may be more difficult to follow along, considering you and I may have differing data. So let's just use a default ggplot2 dataset? Let's work with the mammals sleep dataset (```msleep```).

```{r Mammals Data }
names(msleep) #let's see what columns are in this data...
```

I wonder how body weight and sleep time are related? Let's use proper ggplot, not just qplot for this. Here we will call the ```ggplot()``` command. This command takes two arguments: data and an aesthetic mapping (for a scatter plot we map the data to x,y coordinates of a point).

```{r First ggplot}
p <- ggplot(data=msleep,aes(x=bodywt,y=sleep_total))
```

Yet, this is not enough to make a plot. In ggplot, we still need a layer (in which we specify a point as our geometric object, mapping x,y from our dataset, but if we had a line geomtry we would need to map other aesthetics such as line width, which could just be a constant).

```{r Adding Layer}
p <- p + layer(geom = "point") + theme_bw(base_size=20)
p
```

The layer function can actually become quite verbose, so most geoms have shortcut functions: ```geom_XXX()```. Furthermore, our plot doesn't reveal much information, perhaps a logarithmic scale on the x-axis would be helpful? This can be done by apply continuous transformation to the x-axis.

```{r ggplot points}
p <- ggplot(data=msleep,aes(x=bodywt,y=sleep_total)) + geom_point() + scale_x_continuous(trans="log10") + theme_bw(base_size=20)
p
```

### Fitting Data

This data looks a lot like there might be a power law involved... Let's try to fit it. We can actually use a linear model to fit since we're rescaling our axis to a logarithmic scale, the data appears to be linear and this is what will be fit. This can be done with the ```stat_smooth``` function in ggplot by specifying the method (linear model) and the data. This function will add the fit and a 95% confidence interval to the plot. This fit isn't very good.

```{r ggplot fitting}
p <- p + stat_smooth(method="lm",data=msleep,formula= y~x)
p
```

While this is very pretty, there is no straightfoward way of extracting the fit coefficients from ggplot. But, fear not for models can be generated in R and coefficients can be extracted from those. In fact, we can use the exact same function as ggplot's stat_smooth method, ```lm()``` which is a least squares regression algorithm.

```{r R fitting}
fit <- lm(msleep$sleep_total ~ log10(msleep$bodywt))
summary(fit)
```

The fit object pretty much stores everything. We can extract the coefficients and use them to add a line to our ggplot so we can compare the ggplot fit to the R fit. Unsurprisingly, they overlap.

```{r R and ggplot fit}
p + geom_abline(intercept=fit$coefficients[1],slope=fit$coefficients[2])
```

If you cannot by any means manipulate your data to into a linear form to take advantage of this, there is also a nonlinear least squares fitting function, ```nls()```.

### Plotting Functions

This appears to be a bit of a pain. I haven't had much time or need to look into this. Just use Wolfram Mathematica.

### Plotting Distributions

It is very simple to plot a distribution in ggplot. Let's practice with a dataset. We can check our loaded datasets with the command ```data()```. For now, let's use Edgar Anderson's Iris Data.

```{r iris data}
head(iris)
```

To plot a histogram instead of a scatterplot, we simply change the geom. Furthermore, we can divide the data. For example, we can fill the histogram with a different color based on the species of iris.

```{r plot iris}
ggplot(data=iris,aes(x=Petal.Length,fill=Species)) + geom_histogram() + theme_bw(base_size=20)
```

It is also possible to plot a "smooth density estimate" or a kernel density estimate (KDE) of the distribution.

```{r kde iris}
ggplot(data=iris,aes(x=Petal.Length,fill=Species)) + geom_density() + theme_bw(base_size=20)
```

### Exporting Plots

```{r save,eval=FALSE}
ggsave("my_plot.png",width=5,height=5)
```